Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-01 documents/ms.tex
updating discussion, part of the conclusion and corner plot

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1102}
\newcommand{\nkoimcq}{275}
% \newcommand{\nkoigood}{222}
% \newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

\title{Inferring stellar rotation periods using Gaussian processes}

@@

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \naigrain\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an AutoCorrelation Function
(ACF) method.
The posterior PDF of the rotation period parameter was sampled using the
affine invariant ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee}
\citep{Foreman-Mackey2013}, and the GP operations were performed using the
{\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, rotation periods are inferred probabilistically, allowing them to
be easily incorporated into follow-on studies; inferring stellar ages or
characterising star-planet interactions for example.
In addition, unlike the sine-fitting periodogram and ACF methods, the
posterior PDF samples allow for proper characterisation of uncertainties.
This method eliminates the need to define a heuristic detection criterion such
as are often used in the literature: a light curve that does not contain a
rotation signal will return the prior PDF.
Furthermore, the ACF method is not applicable to unevenly spaced data,
therefore it is inappropriate for the ~$10^7$ stellar light curves expected
to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The RMS scatter about the $x=y$ line for rotation periods measured using the
ACF method is \acfRMS.
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are similar to the QP kernel function in equation
\ref{eq:QP}: periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The RMS scatter for rotation periods measured using the LS periodogram
technique is \pgramRMS.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

@@

In order to manage the flexibilty of the GP model and focus on reliably
retrieving the period parameter, we also impose priors on the non-period GP
parameters, learned during the development of this method.
Beginning with very broad priors (uniform in log space between -20 and 20) on
all the parameters, we first explored the typical values of the non-period
parameters when the injected periods were successfully recovered.
We also experimented with constraining the allowed ranges of the parameters
after discovering that some regions of parameter space (such as large values
of $A$ and $\Gamma$ and small values of $l$) tended to allow fits that ignored
the desired periodicity.
The priors and bounds we used for this work are listed in table
\ref{tab:priors}.
We note that we do not have any specific quantitative justification for the
use of these priors, except that they produce good results for this particular
test set of light curves.
We thus caution that other data sets might have different noise properties,
potentially requiring modifications to these priors on the non-period
parameters (e.g., see Section \ref{sec:kepler}).
For the period prior, we tried two different methods: an uninformed (log-flat)
prior between 0.5\,d and 100\,d, and an ACF-informed prior (details in
Appendix.)
% We note that our default prior for period is log-uniform between the bounds,
% but that we were also able to further improve our results by using an
% ACF-inspired prior.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_24_0.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
    }
\label{fig:compare_mcmc}
\end{center}
\end{figure*}

Figures \ref{fig:compare_mcmc_noprior} and \ref{fig:compare_mcmc_acfprior}
summarize the results of this MCMC fitting
procedure compared to the injected ``true'' stellar rotation periods, for the
uninformed and ACF-informed priors on period, respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/0002_triangle.png}
\caption{(Place holder - to be updated.
Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
The true rotation period of this star is ... ($\exp(...)$) days.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to the
ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping KOIs.
This comparison shows generally very good agreement, with only a few exceptions,
demonstrating that this method works not only on simulated data, but also on
real data---with the caveat that for any particular data set, some care is needed
regarding the setting the priors and ranges for the GP hyperparameters.
>>>>>>> 265508362e5e7a01a39bdd616c754c276059397c

@@

The main drawback of the GP method is computation time.
Because GPs are so expensive to compute, it is necessary to come up with
shortcuts in order to perform inference on \kepler\ light curves, each of
which compromises accuracy.
These shortcuts may only be necessary when working with a large number of
stars (of order hundreds or thousands).
If only interested in single or small numbers of targets it may be possible to
reduce the extremity of these shortcuts or to avoid them altogether.
The shortcuts that we employ here include:
\begin{itemize}
\item{Initialisation.
The choice of initial parameter values makes a difference to computation time:
the closer they are to the maximum likelihood parameters the shorter the MCMC
burn-in time.
Using the ACF period to initialise is not ideal as results will not be fully
independent of the ACF period unless MCMC chains are run for an infinitely
long time.}
\item{Subsampling.
Not only is some information always lost by subsampling, but the choice of
subsampling frequency will also effect the results.
We choose to retain an average of only 4 data points per day, based on the
range of rotation periods we are interested in.}
\item{Priors.
We choose to place a Gaussian prior in log-space over the rotation period
parameter where the standard deviation is twice the log of the ACF rotation
period.
Unfortunately however, if the true period is far different to the ACF period,
the MCMC chains will take a long time to reach the correct value.
In practice we found that the introduction of a Gaussian prior improved
results compared with a broad prior that was uniform in log-space.}
\end{itemize}

\subsection{Initialisation}
Using the ACF to initialise the MCMC chains is not ideal because, of course,
one becomes reliant on the assumption that the ACF period is close to the true
period.
Of course, if you had infinite CPU time, this would not be a problem as you
would eventually sample the entire posterior PDF of the period parameter,
however in practice this is likely to be an issue.
The only way to get around this problem is to run the MCMC chains for as long
as possible, or, alternatively to use a sampler that is designed to move
around the parameter space much more quickly than {\tt emcee}.
% We implemented a version of our code using nested sampling but the results
% were not as good as the {\tt emcee} method.
% We believe this to be due to the initialisation --- there is no initialisation
% in nested sampling, samples are drawn directly from the prior.
% In a problem with a multimodal posterior PDF such as inferring periods,
% initialising MCMC chains close to the maximum likelihood value may lead to a
% more rapid convergence.
We chose to use the ACF periods rather than the periodogram periods to
initialise the results since there were fewer large outliers.

\subsection{Kernel function choice and interpretation of hyper-parameters}
The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
It may in fact be even {\it more} suited to modelling stellar
light curves as it describes a Gaussian in frequency space.
It is easy to imagine a differentially rotating star with a period that is a
Gaussian in frequency space: the mean frequency would be the frequency at the
most active latitude, at or near which spots spend the majority of their time
and the tails would be occupied by spots that drift near the equator or poles.
The main difference between this cosine and the QP function is that the cosine
function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
In other words, we do not want to force anti-correlation of points that are
$\frac{1}{2}$ a period apart.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.

@@

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data and is, in most cases robust to the
presents of transits and eclipses in the light curves.

Although rotation period uncertainties calculated using the GP method are
better estimates than those calculated using the ACF method, they are still
imperfect estimates of the intrinsic uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps, we
would the entire posterior PDF would eventually be sampled, producing an
accurate uncertainty measurement.
Practically speaking however, these imperfect estimates may be the best
currently achievable.

A GP is a semi-parametric model that is well suited to signals that are
non-sinusoidal and quasi-periodic.
Not only this, unlike the other two methods, the GP method provides posterior
PDF samples of rotation periods, allowing characterisation of uncertainties.
Note that the posterior PDFs are often non-Gaussian, being bi- or multimodal.
In these cases it is therefore misleading to report median values or 16th and
84th percentiles as uncertainties.
Rather, it is better to incorporate posterior PDF samples into analyses which
require uncertainty propagation.
Because rotation periods are inferred probabilistically with the GP method,
this allows them to be easily incorporated into follow-on studies; inferring
stellar ages or characterising star-planet interactions for example.

The main drawback of the GP method is the computational cost.

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1102}
\newcommand{\nkoimcq}{275}
% \newcommand{\nkoigood}{222}
% \newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{0.60}
\newcommand{\pgramRMS}{0.98}
\newcommand{\mcmcRMS}{0.46}

@@

\title{Inferring probabilistic stellar rotation periods using Gaussian
processes}

@@

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We use \naigrain\ simulated light curves with a range of rotation periods to
    test the GP model and compare our results with a sine-fitting periodogram
    method and an AutoCorrelation Function (ACF) method.
The posterior PDFs of the parameters were sampled using the affine invariant
    ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee3}
    \citep{Foreman-Mackey2013}, and the GP operations were performed using the
    {\tt george} python package \citep{George}.
We demonstrate that rotation periods inferred via this method are more
    accurate than the periodogram and ACF methods.
Furthermore, the rotation periods are inferred probabilistically, allowing
    them to be easily incorporated into follow-on studies; inferring stellar
    ages or characterising star-planet interactions for example.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The Median Absolute Deviation (MAD) of the ACF-recovered periods is \acfRMS\
in log-space (1.8 days).
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are similar to the QP kernel function in equation
\ref{eq:QP}: periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MAD for rotation periods measured using the LS periodogram technique is
\pgramRMS\ in log-space (2.7 days).
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

@@

In order to manage the flexibilty of the GP model and focus on reliably
retrieving the period parameter, we also impose priors on the non-period GP
parameters, learned during the development of this method.
Beginning with very broad priors (uniform in log space between -20 and 20) on
all the parameters, we first explored the typical values of the non-period
parameters when the injected periods were successfully recovered.
We also experimented with constraining the allowed ranges of the parameters
after discovering that some regions of parameter space (such as large values
of $A$ and $\Gamma$ and small values of $l$) tended to allow fits that ignored
the desired periodicity.
The priors and bounds we used for this work are listed in table
\ref{tab:priors}.
We note that we do not have any specific quantitative justification for the
use of these priors, except that they produce good results for this particular
test set of light curves.
We thus caution that other data sets might have different noise properties,
potentially requiring modifications to these priors on the non-period
parameters (e.g., see Section \ref{sec:kepler}).
For the period prior, we tried two different methods: an uninformed (log-flat)
prior between 0.5\,d and 100\,d, and an ACF-informed prior (details in
Appendix).
% We note that our default prior for period is log-uniform between the bounds,
% but that we were also able to further improve our results by using an
% ACF-inspired prior.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_24_0.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in the appendix.
The MAD of these results is 0.38 in log-space, \ie\ 1.46 days.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_noprior.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with no .
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
An uninformative prior, flat in the natural log of the rotation period between
    0.5 and 100 days was used to generate these results.
The MAD of these results is 0.48 in log-space, \ie\ 1.61 days.
    }
\label{fig:compare_mcmc_noprior}
\end{center}
\end{figure*}

Figures \ref{fig:compare_mcmc_noprior} and \ref{fig:compare_mcmc_acfprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
Both the informative and uninformative prior versions of the GP method perform
better than the ACF and periodogram methods, with MADs of 0.38 and 0.48,
respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/25.png}
\caption{Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
    The green line in the period panel shows the injected period.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to the
ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping KOIs.
This comparison shows generally very good agreement, with only a few exceptions,
demonstrating that this method works not only on simulated data, but also on
real data---with the caveat that for any particular data set, some care is needed
regarding the setting the priors and ranges for the GP hyperparameters.

@@

The uncertainties produced by the GP method are more representative than those
produced by the ACF and periodogram methods because the posterior PDF of the
period is explored using MCMC.
Whilst they are an improvement, unfortunately these uncertainties are still
underestimated in the majority cases.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lay within 1$\sigma$ of the true period.
50\% lay within 2$\sigma$ and 66\% within 3$\sigma$.
The largest outlier was 114$\sigma$ away from the true value.
In the majority of these cases the uncertainties are underestimated due to the
multi-modal nature of the period posterior PDF.
Because the posteriors are multi-modal they are difficult to sample and,
without running the MCMC for infinitely long, the samples will not necessarily
always cover the full range of periods with significant power.
The underestimated uncertainties can also be attributed to the model choice.
Although a GP model is more appropriate than a sinusoid, it is still an
imperfect, effective model.
A perfect physical model of the star would produce better results with more
representative uncertainties.

The QP kernel function represents a simplistic effective model of a stellar
light curve.
It adequately describes the data, captures that all-important periodic quality
and is relatively simple, with only a few hyper-parameters.
It also satisfies the requirement to produce positive semi-definite covariance
matrices.
Whilst the QP kernel function evidently captures the periodic qualities of
light curves adequately, it is still a somewhat arbitrary choice.
Another valid choice would be a squared cosine function multiplied by a
squared exponential,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2}\right)
\cos\left(\frac{2\pi}{P}\right)
\end{equation}
\label{eq:cos_kernel}
This function produces a positive semi-definite matrix and has the $P$
parameter of interest.
It may in fact be even {\it more} suited to modelling stellar
light curves as it describes a Gaussian in frequency space.
It is easy to imagine a differentially rotating star with a period that is a
Gaussian in frequency space: the mean frequency would be the frequency at the
most active latitude, at or near which spots spend the majority of their time
and the tails would be occupied by spots that drift near the equator or poles.
The main difference between this cosine and the QP function is that the cosine
function allows negative covariances and the QP function does not.
Is is realistic to allow negative covariances?
In practice, the ACFs of \Kepler\ light curves often go negative.
However, many stars have two active regions on opposite hemispheres that
produce two brightness dips per rotation.
If the covariance is forced to be negative for two data points that are
separated by half a rotation period, those light curves with two peaks per
rotation period may not be well modelled.
In other words, we do not want to force anti-correlation of points that are
$\frac{1}{2}$ a period apart.
It would be very worthwhile to test this assumption and this alternative
kernel function in future.
If CPU time were not limited there may be some benefit to performing formal
model comparison with different kernel functions.
However, the evidence integral is an ambitious calculation for models with
likelihood functions that take milliseconds to compute, let alone those
involving GPs with light curves containing thousands of data points which can
take minutes.

@@

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
As expected, it provides a large improvement over the periodogram method and a
moderate one over the ACF method.
The periodogram method performs poorly because a sinusoid is not a perfect
model choice for stellar light curves --- they do not always display
sinusoidal behaviour.
On the other hand, the ACF method does not rely on any model choice.

In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

The GP method provides uncertainties because it produces posterior PDF samples
but they are still imperfect estimates of the intrinsic uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, producing an accurate
uncertainty measurement.
Practically speaking however, these imperfect estimates may be the best
currently achievable.

A GP is a semi-parametric model that is well suited to signals that are
non-sinusoidal and quasi-periodic.
Not only this, unlike the other two methods, the GP method provides posterior
PDF samples of rotation periods, allowing characterisation of uncertainties.
The posterior PDFs are often non-Gaussian, being bi- or multimodal.
Rather, it is better to incorporate posterior PDF samples into analyses which
require uncertainty propagation.
Because rotation periods are inferred probabilistically with the GP method,
this allows them to be easily incorporated into follow-on studies; inferring
stellar ages or characterising star-planet interactions for example.
